Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-09 Manuscript/manuscript.tex
New version of density plot and updates on text (I am still working on it)

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{JOE: list more introductory examples, where it has interesting effects.}
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} 
\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
series, the molecular details of binding and binding energetics are
not fully understood \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding models and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~divalent cation-induced charge inversion of bacterial membranes~\cite{Luan2016} \todo{JOE: this may be a too daring statement.}).

@@

The main goal of this work is to apply ECC correction to accurately 
describe the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
in MD simulation models when compared with NMR data \cite{catte16}. 
The Lipid14~\cite{dickson14} force field parameters were used as a starting point,
because their response to bound ions was most realistic against NMR data
(see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for molecules, i.e. lipid head groups in our case. 
Unlike the total charge of cations and anions, atomic partial charges used in
classical MD simulations are not experimental observables and 
can in principle adopt arbitrary values. 
In cases where the partial charges have some physical meaning 
(e.g. approximate electrostatic potential as in RESP~\cite{RESP_paper}),
we can merely estimate the scaling factor for molecules, $f_q$, 
to adopt a value between $\epsilon _{el} ^{-1/2}$ and $1$.
Searching in this parameter space, we empirically discovered that
the scaling factor $f_q = 0.8$ is optimal for applying ECC on Lipid14 model. 
Such a finding is in line with the estimate considering the
so called ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}.
IPolQ charges are obtained as the average of partial charges given by 
restrained electrostatic potential fitting (RESP~\cite{RESP_paper}) 
in vacuum and in a solvent.
Using the partial atomic charges obtained from RESP fitting in vacuum and implicit solvent reported in \cite{maciejewski14} and
considering that the partial charges in original Lipid14 parametrization were derived using RESP~\cite{RESP_paper} in vacuum,
we find out that the charges in Lipid14 model shall be scaled roughly by the factor $f_q=0.8$ 
-- agreeing with our empirical finding. 
%By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
%we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
%by increasing the scaling factor $f_q$ to $0.8$. 
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

Here we apply the ECC correction with the proposed scaling factor only to the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized in most lipids models including Lipid14 and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16}.
In contrast, all available lipid models require
some improvements in glycerol backbone and head group regions \cite{botan15}. 
%To minimize the detuning of the highly optimized hydrocarbon
%parameters, 
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications.

The scaling of partial charges in the polar region was found to 
significantly decrease cation binding and
the area per molecule to $\approx$60~\AA$^2$, which is significantly smaller than the
experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment}) and the original Lipid14 values $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
%\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
%JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. 
The solvation free energy and hydration is sensitive to Van der Waals interaction parameters,
hence, they can be increased by reducing the radius of the selected atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential. 
%This allows water molecules to approach closer to lipid atoms and
%have stonger electrostatic interactions with them.
%\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Here we reduce $\sigma$ with a scaling factor of $f_\sigma = 0.89$ for the same atoms for which
charges were scaled. This has lead to an increase of the area per molecule back
close to the experimental value (see Table~\ref{tab:apls}). 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for C-H bonds can be accurately measured for each
C-H bond of lipids by using H$^2$ NMR or $^{13}$C NMR techniques and
calculated from MD simulations directly from definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Combination of atomic absorption spectra, $^2$H NMR experiments
and information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}. However,
it also turned out that the sensitivity of the order parameters response
to the bound charge has to be quantified more accurately for quantitative comparison
of binding affinity. Hence, we first quantify the response of
head group order parameters against experimental data from cationinc surfactants
embedded in PC bilayer~\cite{scherer89}. Essentially all surfactants locate in
lipid bilayers in these experiments, thus the amount of the bound charge is known
and can be plotted against order parameter change.
Secondly we evaluate the response of the head group order parameters to varying salt concentrations in solution
and estimate the amount of membrane bound charge.

%\todo{ongoing,Actual concentration of cations in simulation has yet to be estimated. 
%If it varies too much from the nominal concentration, I may need to tweak the scaling 
%factors, $f_q$ or only $f_\sigma$, to accomodate it. However, it is very unlikely, response 
%to the surfactant \ce{DHMDMAB} is OK. Big patches with loads of solvent are running at the 
%moment to guide me on the possible finite-size errors and this conc-error.
The used experimental data report order parameters as a function of
equilibrium cation concentration in the bulk solvent \cite{akutsu81,altenbach84}.
Such a condition is reached in simulations by adjusting
the simulation box size to dimensions large enough
that ion conceration reaches a clear plateau in the bulk solvent.
The concentrations in the units of $\mathrm{mol/l}$ were then 
determined as
\begin{equation}
\label{eq:conc}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in the units of $\mathrm{nm}^{-3}$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
These systems were ran with original Lipid14 POPC model and with ECC-corrected model.
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}

@@

Headgroup order parameter response to the bound charge was evaluated against
experimental data measured with the cationic surfactant
(dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) \cite{scherer89}.
The exact amount of bound charge in the membrane is known in these systems, because
practically all cationic surfactant molecules are embedded in the lipid bilayer due to
their amphiphilic nature. Thus, the changes of head group order parameter as a function of
mole fraction of cationic surfactants (Fig. \ref{OrderParameterCHANGESnewMODELS}) gives
also the order parameter changes as a function of the amount of bound cations.
\todo{Split the OP-changes figure in two: cationic surfactant -and- binding of cations.
      The separate figure with the cationic surfactant shall be additionally accompanied
      with a plot of Head group tilt (PN-vector?) change/response. }

The head group order parameter response to the bound cation concentration
is in experiments approximately linear up to $\sim$0.3 mole fraction of the cation \cite{scherer89}.
The linearity is also observed in simulations with original Lipid14 and with newly derived ECC-lipids models.
Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
of the response of Lipid14 model is found to be slightly larger than in 
experiments (VALexp vs VALsim
\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
  SAMULI: Maybe these could be put in the same figure with the order parameter
  response and P-N vector angle change with surfactants.
}
), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of head group order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the head group order parameter response
to the bound charge.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities.eps}
  \caption{\label{fig:cacl-dens}
    Profiles of normalized number density of \ce{Ca^{2+}} and \ce{Cl^-} for Lipid14 model with Dang ions
    and with ECC-ions and ECC-lipids model with ECC-ions. 
    Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}), 2 for divalent ions (i.e.~\ce{Ca^2+}), and $1/100$ for water. 
    }
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{What are the concentrations (calculated as in NMRlipids II) in this figure?} \\
  \todo{These should be also compared to the results with Åqvist ions, which are shown
    in NMRlipids II. NMRlipids II have results with 350mM and 1000mM. If we have the distributions with
    the same concentrations Åqvist results could be added also in this figure. If not, the
  comparison has to be made some other way.}
\end{figure}

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{JOE: list more introductory examples, where it has interesting effects.}
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} 
\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
series, the molecular details of binding and binding energetics are
not fully understood \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding models and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~divalent cation-induced charge inversion of bacterial membranes~\cite{Luan2016}
\todo{JOE: this may be a too daring statement. \\
  SAMULI: I think that the divalent cation induced charge inversion is not an artefact,
  but it might happen with too small concentrations. 
  On the other hand, it is not a new observation either, see e.g. literature cited in NMRlipids IV project.}).

@@

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e. lipids in our case. Unlike the total charge of ions, atomic partial charges in
classical MD simulations are not experimental observables.
Instead they are based on theoretical calculations and the
derived values depend, for example, on the conformational
sampling of the molecules and description of solvent 
in the calculations \cite{RESP_paper,??}. 
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~\cite{RESP_paper}),
Depending to the details, partial charge calculations may partially
incude some of the solvent electronic polarizability effects. Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges would lie
between $\epsilon _{el} ^{-1/2}$ (no electronic polarizability in calculation)
and $1$ (all electronic polarizability in calculation).

In this work, we empirically sample this parameter space
to find a atomistic MD simulation force field parameters for lipids that
accurately describes the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
when compared with NMR data \cite{catte16}. This data can used to accurately asses
ion binding in PC bilayers, as discussed in section \ref{section:electrometer}.
We chose the Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, applied to headgroup, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related headgroup order parameter response became weaker with the scaling of partial
charges. The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
\todo{SAMULI: I am still not sure if this is correct and/or makes sense.}
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters of the same atoms for which charges were scaled
with a scaling factor of $f_\sigma = 0.89$, the area per molecule value was
againg in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for all C-H bonds in lipid molecules
can be accurately measured by using H$^2$ NMR or $^{13}$C NMR techniques.
The same order parameters can be calculated from MD simulations directly
from the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data, combined with the
information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}.
However, only systems with lipid bilayers and aqueous cations were used in that
study. The headgroup response to bound charge then depends on both, the
binding affinity and the response of the headgroup on bound charge.
Thus, the headgroup response to bound charge has to be correctly
described in the model if the concept is used for quantitative studies of
binding affinity. Hence, we first quantify the response of head group order
parameters to the bound charge by using experimental data measured from mixtures of cationinc surfactants
and PC lipids~\cite{scherer89}. The amount of the bound charge is known excatly in
these systems, because essentially all surfactants with unit charge locate in
lipid bilayers. Thus, the headgroup order parameter response to bound charge can be first
quantified aginst experimental data before proceeding to binding affinity studies.
%Secondly we evaluate the response of the head group order parameters
%to varying salt concentrations in solution
%and estimate the amount of membrane bound charge.

The experimental data used to quantify cation binding reports
order parameters as a function of equilibrium cation concentration
in the bulk solvent \cite{akutsu81,altenbach84}. Recent work
by NMRlipids project estimated this concentration by using
the molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M,\\
\todo{SAMULI: I call here concentration we used to call ''nominal'' as ''molar concentration in water''.
  I think that the ''nominal'' is not a good term for this. We need to decide how we will call this.} \\
which was accurate enough for the conclusions in that study \cite{catte16}.
Since we are now targeting a model with quantitatively accurate binding
affinity, we used larger simulation box dimensions that allow the
determination of bulk ion conceration. Equilibrium concentration
corresponding the experimental data in the units of $\mathrm{mol/l}$
can be then determined from the ion concentration plateau in the bulk s
olvent at the boundaries of the simulation box as
\begin{equation}
\label{eq:conc}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in the units of $\mathrm{nm}^{-3}$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
These systems were ran with original Lipid14 POPC model and with ECC-corrected model.
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}

@@

Headgroup order parameter response to the amount of bound charge can
be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
All surfactants locate in lipid bilayers due to their amphiphilic nature,
thus the exact amount of bound charge in the membrane is known in these systems.
The headgroup order parameter changes as a function of bound charges can be then
explicitly determined from such systems.

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}).
The head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids.

of the cation \cite{scherer89}.
The linearity is also observed in simulations with original Lipid14 and with
newly derived ECC-lipids models \todo{Yet to be shown.}.
Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
%Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
%of the response of Lipid14 model is found to be slightly larger than in 
%experiments (VALexp vs VALsim
%\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
%  SAMULI: Maybe these could be put in the same figure with the order parameter
%  response and P-N vector angle change with surfactants.
%}
%), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of head group order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the head group order parameter response
to the bound charge.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with Aqvist ions are taken directly from Ref. \citenum{catte16}
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-}) and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration in water is 350~mM in all systems
    presented here. 
    }
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{Resuls with Lipid14+Dang to be added once done.}
\end{figure}
